Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-08-01 pirouette_supplement.tex
Remove 'new' tags, fix #122

\begin{itemize}
  \item{
    subsection \ref{subsec:guidelines}: guidelines for users
  }
  \item{
    subsection \ref{subsec:installation}: installation
  }
  \item{
    subsection \ref{subsec:resources}: resources, such as 
    website, tutorials, packages used, bug reporting and contributing
  }
  \item{
    subsection \ref{subsec:citation}: citation of pirouette
  }
  \item{
    subsection \ref{subsec:twinning}: the
    twinning process
  }
  \item{
    subsection \ref{subsec:candidates}: candidate models for the inference
  }
  \item{
    subsection \ref{subsec:stochasticity}: the effects of stochasticity
  }
  \item{
    subsection \ref{subsec:nltt}: the
    nLTT statistic
  }
  \item{
    subsection \ref{subsec:main_functions}: main functions
  }
  \item{
    subsection \ref{subsec:main_example}: code, extra figures and
    diagnostics regarding the main example.
  }
  \item{
    subsection \ref{subsec:distribution}: the result of using 
    multiple trees, as generated by the
    same stochastic process as the main example
  }
  \item{
    subsection \ref{subsec:n_taxa}: the effect of the number of taxa
  }
  \item{
    subsection \ref{subsec:n_nucleotides}: the effect of the DNA
    alignment sequence length
  }
  \item{
    subsection \ref{subsec:simplest_correct_parameterization} shows the
    effect when performing inference in the simplest use case
  }
  \item{
    subsection \ref{subsec:under_parameterization} shows the
    effect when performing inference with an under-parameterization
  }
  \item{
    subsection \ref{subsec:different_n_mutations} shows the
    effect when the twin alignment is allowed to have a different
    number of \new{substitution}s
  }
  \item{
    subsection \ref{subsec:mutation_rate} shows the
    effect of different mutation rates
  }
  \item{
    subsection \ref{subsec:acknowledgments}: Acknowledgments
  }
  \item{
    subsection \ref{subsec:data_accessibility}: Data accessibility
  }
  \item{
    subsection \ref{subsec:author_contributions}: Author contributions
  }
\end{itemize}

@@

\begin{itemize}
  \item{
    The use of 20 replicates results in decent plots.
  }
  \item{
    The use of more taxa increases the inference error
  }
  \item{
    The use of longer DNA sequences decreases the inference error.
  }
  \item{
    When we do not impose the same number of \new{substitution}s 
    between true and twin alignment, we observe a difference in the error 
    distributions with respect to the standard case (presented in the main 
    text) where they are forced to have the same number of \new{substitution}s.
  }
  \item{
    Using a mutation rate less than 1.0 / crown age, decreases the
    inference error. We predict this will increase the error in the parameter
    estimation.
  }
\end{itemize}

@@

\new{
  We chose to preserve the original topology to 
  increase the similarity between the twin to the original tree. 
  This works well in the cases of BD or DD models 
  we consider in our example, because all these models make the same
  assumption about topology (all topologies are equally likely).
  However, this might not be suitable for new models that assign different probabilities
  to trees with the same branching times but different topologies.
}
\new{
  The default option for the twin diversification model $p_T$ 
  is the standard BD model.
}
\verb;pirouette;
\new{
   has a built-in function to use a Yule model as well.
  Additionally, a user can specify a function to generate a twin tree
  from any speciation model, such as, for example, a coalescent model.
}

It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find 
the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a BD model) 
that maximize this likelihood applied to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology. 
We simulate the new branching times using the TESS 
package [\cite{TESS}].
\new{
  For simplicity, when simulating phylogenies we assumed a sampling fraction 
  of $100\%$. A different choice might have an effect on model performance.
}

The second step ($T_2$) of the twinning process simulates the twin alignment 
with the same clock model, site model and mutation rate 
used to simulate the alignment on the true. 
The twin alignment can be simulated in any user-defined way.
\verb;pirouette; provides the option simulate it with the same mutation rate as the true alignment. By default, however,
not only the same mutation rate is used, but also the total number of 
\new{substitution}s
matches the true alignment. The total number of \new{substitution}s is defined
as the number of different nucleotides between the (known) root sequence
compared to the sequences at the tips.

@@

The goal is to evaluate BEAST2's performance 
on a non-standard tree model, 
one must also consider the last source of stochasticity: 
the different phylogenies a tree model generates.
A single phylogeny cannot be considered as fully representative of the model. 
For this reason multiple phylogenies must be 
considered \new{(at least 100 independent true and twin trees)}.
If the number of considered phylogenies is high enough, 
the comparison between the main pipeline's aggregated error distribution 
and its twin counterpart leads to a fair evaluation 
of the new tree model with respect to the baseline error.

@@

The nLTT statistic is the absolute difference
between the normalized lineages-through-time plots of two trees.
The nLTT statistic is chosen, as it can operate on any two trees (regardless
of their crown ages and number of taxa) and its results have a clear range
from zero to one. This normalized result makes it possible to compare trees 
from a distribution of trees from any tree model.
\new{
  The nLTT statistic is not suitable, however, to distinguish
  between a constant-rate BD model and a family of time-dependent 
  models [\cite{louca2020extant}].
}

@@

The pipeline starts at the top-left panel of figure \ref{fig:pipeline} (which 
is identical to figure \ref{fig:dd_tree}),
which is the 'true tree'.
The 'true tree' is generated by the diversity-dependent (DD) 
tree model [\citep{DDD, etienne2012diversity}],
which is a BD model with a speciation rate that is dependent on the number of species,
with (an arbitrarily chosen) crown age of 10 time units 
and an expected number of 6 tips for an extinction rate of 0.1. 
The carrying-capacity is set to 6. 
The initial speciation rate $\lambda_0$ is chosen 
such that the expected number of species in a constant-rate BD model 
would be equal to the number of tips, which amounts to $\lambda_0 = 0.63$.
\new{
  Note that in the main example, a tree was generated with 5 tips, due
  to stochasticity in the tree generation algorithm.
}

@@

Based on the 'twin alignment' (center-right panel), the candidate model
with the highest marginal likelihood is determined, from the same 
set of 15 candidate models. The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_twin}) was 
the inference model with a JC nucleotide substitution model,
an RLN clock model and a Yule tree model 
\new{
  (Note that this is different from how the twin tree was generated which was with a BD process and the alignment
  was simulated with a JC substitution model and strict clock model
  ). However, the extinction rate used in simulating the twin tree was practically 0, thus resembling a Yule process.
}
From the 'twin alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the fourth panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

\new{
  The resulting error distributions are shown in Fig. \ref{fig:replicate_trees}. 
  We present results for cases where (1) the generative model has been used 
  or (2) the model with highest evidence has been selected for the inference.
  From the plots we can see that in both cases the two distributions 
  (true and twin) are mostly overlapping, but not everywhere.
  This suggests that the inference 
  models that have been used can to a reasonable extent capture in an accurate way 
  the features of the diversity-dependent tree prior used to simulate 
  the original trees. 
}

@@

\new{
  We show in figures \ref{fig:replicate_trees}, \ref{fig:example_12_taxa}, 
  \ref{fig:example_24_taxa}, \ref{fig:example_32_taxa} and 
  \ref{fig:example_40_taxa} what are the errors obtained when starting from 
  phylogenies with, respectively, 5, 12, 24, 32 and 40 taxa.
  Again we can see 
  that in each case errors tend to be greater in the true distribution than 
  in the twin distribution, similar to the result of subsection \ref{subsec:distribution}.
  Collecting all the data together we can see that errors tend to decrease as the number of taxa in the considered phylogenies increase (see Fig. \ref{fig:error_vs_ntaxa}). The data point for $40$ taxa not following the trend could be due to the limited amount of simulated trees taken in consideration due to time constraints.
}

@@

\new{
  From figures \ref{fig:example_500_nucleotides}, 
  \ref{fig:example_1000_nucleotides} and \ref{fig:example_2000_nucleotides} 
  we can observe that the discrepancy between the true and twin error 
  distributions tends to become smaller as the number of nucleotides increase (see also Fig. \ref{fig:error_vs_seqlength}).
  This occurred for both the generative and best candidate cases.
  This follows the expectation that a prior becomes less important when
  more information becomes available.
}

@@

\new{
  As expected the twin and true distributions in Fig. \ref{fig:example_yule} 
  are extremely similar for both the generative and the best candidate case.
}

@@

\new{
  Because the two models are very similar to each other (the BD model can be 
  turned into a Yule model just by setting the extinction parameter to 
  zero [\cite{nee1994reconstructed}]) the median discrepancy is
  almost negligible. 
  However, with respect to the previous 
  case (subsection \ref{subsec:simplest_correct_parameterization}), 
  where a Yule tree prior was used, 
  the distributions here exhibit a greater difference.
}
\new{As we use only extant trees, it is reasonable that the method
  is slightly weaker in distinguishing between the Yule and BD
  models. It is unknown what the discriminatory power would be
  when comparing trees with extinction events.}

@@

\new{
  Here the median errors are similar in the three settings and similar to 
  ones relative to the full dataset of \ref{subsec:distribution}. We can also 
  notice that in the case of median likelihood, the twin median error appears 
  to be lower than the true mean error. This is usually a sign that the number 
  of replicates (in this case 10) is too low to allow us to draw precise 
  conclusions from this test. We did not explore further in this 
  direction using more simulations because computational times turned to be 
  extremely high. 
  The entire run took 120 hours in total.
}

@@

The main example uses a twin alignment that has the same number
of \new{substitution}s (as measured from the ancestral sequence) as the true alignment. Here, we show the same results, with the difference that
the twin alignment uses the same mutation rate, yet is not guaranteed
to have the same number of \new{substitution}s.

\begin{figure}[H]
  %\includegraphics[width=0.98\textwidth]{pirouette_example_18/errors.png}
  \includegraphics[width=0.94\textwidth]{pirouette_example_18/errors.png}
  \caption{Aggregate error distributions for 100 replicates.
    similar to Fig. \ref{fig:replicate_trees}, 
    but here the number of \new{substitution}s is not imposed 
    to be the same between true and twin alignment. 
    Instead, an equal mutation rate is used. 
    This took 3.3 days (wall clock time) to compute.}
  \label{fig:example_random_mutations}
\end{figure}

\new{
  Comparing figures \ref{fig:example_random_mutations} 
  and \ref{fig:replicate_trees} we can see that the discrepancy between 
  true and twin distributions tend to increase. 
  This is probably due to the fact that letting mutation rates 
  induces a difference in the amount of information contained in the 
  alignments and this is reflected in the error distributions.
}

@@

Here, we show the same results for different mutation rates.
\new{
\iffalse
  We can observe that the error tends to increases for mutation rates above
  `1.0 / crown age`. Below this value, the errors are the same.
  For these lower mutation rates, however, the shape of the error
  distribution changes for the baseline/generative models,
  deviating more and more from the gamma-distribution-like shape
  as is still present in the twin distributions.
\fi
 The results for the different mutation rates are shown in Figs. \ref{fig:example_0.25_mutation_rate} (0.25 / crown age), \ref{fig:example_0.50_mutation_rate} (0.50 / crown age), \ref{fig:example_0.75_mutation_rate} (0.75 / crown age), \ref{fig:replicate_trees} (1.00 / crown age), \ref{fig:example_1.25_mutation_rate} (1.25 / crown age),
 \ref{fig:example_1.50_mutation_rate} (1.50 / crown age) and
 \ref{fig:example_2.00_mutation_rate} (2.00 / crown age).
 }
\new{
 Fig. \ref{fig:error_vs_mutationrate} summarizes all the other figures showing on the y-axis, for each value of the mutation rate, the difference between the median of the true distribution and the median of the twin distribution. We can observe a general positive trend as the mutation rate increase, even though the value for 1.5 / crown age suggests to take this result with caution. It is possible, however, that a more regular trend could be observed increasing the number of simulations.
}

>>>>

\begin{itemize}
  \item{
    subsection \ref{subsec:guidelines}: guidelines for users
  }
  \item{
    subsection \ref{subsec:installation}: installation
  }
  \item{
    subsection \ref{subsec:resources}: resources, such as 
    website, tutorials, packages used, bug reporting and contributing
  }
  \item{
    subsection \ref{subsec:citation}: citation of pirouette
  }
  \item{
    subsection \ref{subsec:twinning}: the
    twinning process
  }
  \item{
    subsection \ref{subsec:candidates}: candidate models for the inference
  }
  \item{
    subsection \ref{subsec:stochasticity}: the effects of stochasticity
  }
  \item{
    subsection \ref{subsec:nltt}: the
    nLTT statistic
  }
  \item{
    subsection \ref{subsec:main_functions}: main functions
  }
  \item{
    subsection \ref{subsec:main_example}: code, extra figures and
    diagnostics regarding the main example.
  }
  \item{
    subsection \ref{subsec:distribution}: the result of using 
    multiple trees, as generated by the
    same stochastic process as the main example
  }
  \item{
    subsection \ref{subsec:n_taxa}: the effect of the number of taxa
  }
  \item{
    subsection \ref{subsec:n_nucleotides}: the effect of the DNA
    alignment sequence length
  }
  \item{
    subsection \ref{subsec:simplest_correct_parameterization} shows the
    effect when performing inference in the simplest use case
  }
  \item{
    subsection \ref{subsec:under_parameterization} shows the
    effect when performing inference with an under-parameterization
  }
  \item{
    subsection \ref{subsec:different_n_mutations} shows the
    effect when the twin alignment is allowed to have a different
    number of substitutions
  }
  \item{
    subsection \ref{subsec:mutation_rate} shows the
    effect of different mutation rates
  }
  \item{
    subsection \ref{subsec:acknowledgments}: Acknowledgments
  }
  \item{
    subsection \ref{subsec:data_accessibility}: Data accessibility
  }
  \item{
    subsection \ref{subsec:author_contributions}: Author contributions
  }
\end{itemize}

@@

\begin{itemize}
  \item{
    The use of 20 replicates results in decent plots.
  }
  \item{
    The use of more taxa increases the inference error
  }
  \item{
    The use of longer DNA sequences decreases the inference error.
  }
  \item{
    When we do not impose the same number of substitutions 
    between true and twin alignment, we observe a difference in the error 
    distributions with respect to the standard case (presented in the main 
    text) where they are forced to have the same number of substitutions.
  }
  \item{
    Using a mutation rate less than 1.0 / crown age, decreases the
    inference error. We predict this will increase the error in the parameter
    estimation.
  }
\end{itemize}

@@

We chose to preserve the original topology to 
increase the similarity between the twin to the original tree. 
This works well in the cases of BD or DD models 
we consider in our example, because all these models make the same
assumption about topology (all topologies are equally likely).
However, this might not be suitable for new models that assign different probabilities
to trees with the same branching times but different topologies.
The default option for the twin diversification model $p_T$ 
is the standard BD model.
\verb;pirouette; has a built-in function to use a Yule model as well.
Additionally, a user can specify a function to generate a twin tree
from any speciation model, such as, for example, a coalescent model.

It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find 
the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a BD model) 
that maximize this likelihood applied to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology. 
We simulate the new branching times using the TESS 
package [\cite{TESS}].
For simplicity, when simulating phylogenies we assumed a sampling fraction 
of $100\%$. A different choice might have an effect on model performance.

The second step ($T_2$) of the twinning process simulates the twin alignment 
with the same clock model, site model and mutation rate 
used to simulate the alignment on the true. 
The twin alignment can be simulated in any user-defined way.
\verb;pirouette; provides the option simulate it with the same mutation rate as the true alignment. By default, however,
not only the same mutation rate is used, but also the total number of 
substitutions
matches the true alignment. The total number of substitutions is defined
as the number of different nucleotides between the (known) root sequence
compared to the sequences at the tips.

@@

The goal is to evaluate BEAST2's performance 
on a non-standard tree model, 
one must also consider the last source of stochasticity: 
the different phylogenies a tree model generates.
A single phylogeny cannot be considered as fully representative of the model. 
For this reason multiple phylogenies must be 
considered (at least 100 independent true and twin trees).
If the number of considered phylogenies is high enough, 
the comparison between the main pipeline's aggregated error distribution 
and its twin counterpart leads to a fair evaluation 
of the new tree model with respect to the baseline error.

@@

The nLTT statistic is the absolute difference
between the normalized lineages-through-time plots of two trees.
The nLTT statistic is chosen, as it can operate on any two trees (regardless
of their crown ages and number of taxa) and its results have a clear range
from zero to one. This normalized result makes it possible to compare trees 
from a distribution of trees from any tree model.
The nLTT statistic is not suitable, however, to distinguish
between a constant-rate BD model and a family of time-dependent 
models [\cite{louca2020extant}].

@@

The pipeline starts at the top-left panel of figure \ref{fig:pipeline} (which 
is identical to figure \ref{fig:dd_tree}),
which is the 'true tree'.
The 'true tree' is generated by the diversity-dependent (DD) 
tree model [\citep{DDD, etienne2012diversity}],
which is a BD model with a speciation rate that is dependent on the number of species,
with (an arbitrarily chosen) crown age of 10 time units 
and an expected number of 6 tips for an extinction rate of 0.1. 
The carrying-capacity is set to 6. 
The initial speciation rate $\lambda_0$ is chosen 
such that the expected number of species in a constant-rate BD model 
would be equal to the number of tips, which amounts to $\lambda_0 = 0.63$.
Note that in the main example, a tree was generated with 5 tips, due
to stochasticity in the tree generation algorithm.

@@

Based on the 'twin alignment' (center-right panel), the candidate model
with the highest marginal likelihood is determined, from the same 
set of 15 candidate models. The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_twin}) was 
the inference model with a JC nucleotide substitution model,
an RLN clock model and a Yule tree model 
(Note that this is different from how the twin tree was generated which was with a BD process and the alignment
was simulated with a JC substitution model and strict clock model). 
However, the extinction rate used in simulating the twin tree was practically 0, thus resembling a Yule process.
From the 'twin alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the fourth panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

The resulting error distributions are shown in Fig. \ref{fig:replicate_trees}. 
We present results for cases where (1) the generative model has been used 
or (2) the model with highest evidence has been selected for the inference.
From the plots we can see that in both cases the two distributions 
(true and twin) are mostly overlapping, but not everywhere.
This suggests that the inference 
models that have been used can to a reasonable extent capture in an accurate way 
the features of the diversity-dependent tree prior used to simulate 
the original trees.

@@

We show in figures \ref{fig:replicate_trees}, \ref{fig:example_12_taxa}, 
\ref{fig:example_24_taxa}, \ref{fig:example_32_taxa} and 
\ref{fig:example_40_taxa} what are the errors obtained when starting from 
phylogenies with, respectively, 5, 12, 24, 32 and 40 taxa.
Again we can see 
that in each case errors tend to be greater in the true distribution than 
in the twin distribution, similar to the result of subsection \ref{subsec:distribution}.
Collecting all the data together we can see that errors tend to decrease as the number of taxa in the considered phylogenies increase (see Fig. \ref{fig:error_vs_ntaxa}). The data point for $40$ taxa not following the trend could be due to the limited amount of simulated trees taken in consideration due to time constraints.

@@

From figures \ref{fig:example_500_nucleotides}, 
\ref{fig:example_1000_nucleotides} and \ref{fig:example_2000_nucleotides} 
we can observe that the discrepancy between the true and twin error 
distributions tends to become smaller as the number of nucleotides increase (see also Fig. \ref{fig:error_vs_seqlength}).
This occurred for both the generative and best candidate cases.
This follows the expectation that a prior becomes less important when
more information becomes available.

@@

As expected the twin and true distributions in Fig. \ref{fig:example_yule} 
are extremely similar for both the generative and the best candidate case.

@@

Because the two models are very similar to each other (the BD model can be 
turned into a Yule model just by setting the extinction parameter to 
zero [\cite{nee1994reconstructed}]) the median discrepancy is
almost negligible. 
However, with respect to the previous 
case (subsection \ref{subsec:simplest_correct_parameterization}), 
where a Yule tree prior was used, 
the distributions here exhibit a greater difference.

As we use only extant trees, it is reasonable that the method
is slightly weaker in distinguishing between the Yule and BD
models. It is unknown what the discriminatory power would be
when comparing trees with extinction events.

@@

Here the median errors are similar in the three settings and similar to 
ones relative to the full dataset of \ref{subsec:distribution}. We can also 
notice that in the case of median likelihood, the twin median error appears 
to be lower than the true mean error. This is usually a sign that the number 
of replicates (in this case 10) is too low to allow us to draw precise 
conclusions from this test. We did not explore further in this 
direction using more simulations because computational times turned to be 
extremely high. 
The entire run took 120 hours in total.

@@

The main example uses a twin alignment that has the same number
of substitutions (as measured from the ancestral sequence) as the true alignment. Here, we show the same results, with the difference that
the twin alignment uses the same mutation rate, yet is not guaranteed
to have the same number of substitutions.

\begin{figure}[H]
  %\includegraphics[width=0.98\textwidth]{pirouette_example_18/errors.png}
  \includegraphics[width=0.94\textwidth]{pirouette_example_18/errors.png}
  \caption{Aggregate error distributions for 100 replicates.
    similar to Fig. \ref{fig:replicate_trees}, 
    but here the number of substitutions is not imposed 
    to be the same between true and twin alignment. 
    Instead, an equal mutation rate is used. 
    This took 3.3 days (wall clock time) to compute.}
  \label{fig:example_random_mutations}
\end{figure}

Comparing figures \ref{fig:example_random_mutations} 
and \ref{fig:replicate_trees} we can see that the discrepancy between 
true and twin distributions tend to increase. 
This is probably due to the fact that letting mutation rates 
induces a difference in the amount of information contained in the 
alignments and this is reflected in the error distributions.

@@

Here, we show the same results for different mutation rates.
 The results for the different mutation rates are shown in Figs. \ref{fig:example_0.25_mutation_rate} (0.25 / crown age), \ref{fig:example_0.50_mutation_rate} (0.50 / crown age), \ref{fig:example_0.75_mutation_rate} (0.75 / crown age), \ref{fig:replicate_trees} (1.00 / crown age), \ref{fig:example_1.25_mutation_rate} (1.25 / crown age),
 \ref{fig:example_1.50_mutation_rate} (1.50 / crown age) and
 \ref{fig:example_2.00_mutation_rate} (2.00 / crown age).
Fig. \ref{fig:error_vs_mutationrate} summarizes all the other figures showing on the y-axis, for each value of the mutation rate, the difference between the median of the true distribution and the median of the twin distribution. We can observe a general positive trend as the mutation rate increase, even though the value for 1.5 / crown age suggests to take this result with caution. It is possible, however, that a more regular trend could be observed increasing the number of simulations.
